Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-01-24 paper/paper.tex
update multifixed point inference rules

To show the strong equivalence of $NFA(g \oplus g')$ and $g \oplus g'$, we will demonstrate an isomorphism between them
with two proof terms in our logic that mutually invert each other. To define a map $NFA(g \oplus g') \to g \oplus g'$
we will use a multi-fold over all the mutually recursive defintions of $NFA(g \oplus g')$. Note, even though multiple
fixed points are not taken as primitive, using Beki\'c's theorem we can define admissible multiple fixed points types. Below we give rules to form $\mathsf{cons}$ and $\mathsf{mfold}$ terms to serve as the introduction and elimination rules, respectively, for multi-term fixed points,

\[
  \inferrule
  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}) . X_{k}}
\]
where $\sigma$ is the substitution that unrolls the mutually recursive definition one level by replacing each $X_{k}$ with the fixed point,
\steven{I'm not sure this is right. TODO sanity check it with a small example}
\begin{align*}
  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}) . X_{1} / X_{1} , \dots, \\
  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}) . X_{n} / X_{n} \}
\end{align*}

\[
  \inferrule{\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
             \dots \\
             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
  }{\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
\]

@@

When defining a derivation from $NFA(g \oplus g')$ to $g \oplus g'$, we
wish to simultaneously fold over the mutually recursive definitions comprising $NFA(g \oplus g')$. We may think of this
as an eliminator that replaces each state inside of $NFA(g \oplus g')$ with a tail-recursive call that will ultimately
return a term of type $g \oplus g'$. The structure of such a derivation is given below,

\[
  \inferrule{
    \Delta \vdash e : NFA(g \oplus g') \\
    y_{s} : \simulsubst {(X_{s}' \oplus Z'_{s})} \gamma \vdash p_{s} : g \oplus g' \\
    x_{s}' : \simulsubst {(\simulsubst {X_{s}} {\delta_{X}})}{\gamma} \vdash p_{x,s} : g \\
    x_{j}' : \simulsubst {(\simulsubst {X_{j}} {\delta_{X}})}{\gamma} \vdash p_{x,j} : X_{j} \text{ where } j \not \in \{s, e\} \\
    x_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{x, e} : X_{e} \\
    z_{s}' : \simulsubst {(\simulsubst {Z_{s}} {\delta_{Z}})}{\gamma} \vdash p_{z,s} : g' \\
    z_{j}' : \simulsubst {(\simulsubst {Z_{j}} {\delta_{Z}})}{\gamma} \vdash p_{z,j} : Z_{j} \text{ where } j \not \in \{s, e\} \\
    z_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{z,e} : Z_{e} \\
    y_{e} : \simulsubst I {\gamma} \vdash p_{e} : I
  }{\Delta \vdash \mathsf{mfold}(e) : g \oplus g'}
\]

@@

where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively.
This finshes the construction of a parse transformer from $NFA(g \oplus g')$ to $g \oplus g'$.
The map in the other direction is given the derivation,

\[
\inferrule{p_{1} : g \vdash \textbf{in}_{X_{s}'}(\phi(p_{1})) : NFA(g \oplus g') \\ p_{2} : g' \vdash \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) : NFA(g \oplus g') \\ p : g \oplus g' \vdash p : g \oplus g'}{p : g \oplus g' \vdash \textbf{case}~p \{\textbf{in}_{g} p_{1} \mapsto \textbf{in}_{X_{s}'}(\phi(p_{1})), \\ \textbf{in}_{g'} p_{2} \mapsto \textbf{in}_{{Z_{s}'}}(\psi(p_{2})) \} : NFA(g \oplus g')}
\]

>>>>

To show the strong equivalence of $NFA(g \oplus g')$ and $g \oplus g'$, we will demonstrate an isomorphism between them
with two proof terms in our logic that mutually invert each other. To define a map from $NFA(g \oplus g')$ to $g \oplus g'$
we will use a multi-fold over all the mutually recursive defintions of $NFA(g \oplus g')$. Note, even though multiple
fixed points are not taken as primitive, using Beki\'c's theorem we can define an admissible axiom schema to define multiple fixed points types. Below we give rules to form $\mathsf{cons}$ and $\mathsf{mfold}$ terms to serve as the introduction and elimination rules, respectively, for multi-term fixed points. First, the introduction rule,

\[
  \inferrule
  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}
\]
where $\sigma$ is the substitution that unrolls the mutually recursive definition one level by replacing each $X_{k}$ with the fixed point $\mu(X_{j} = A_{j}).X_{k}$,
\steven{I'm not sure the cons rule is right. TODO sanity check it with a small example}

\begin{align*}
  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \\
  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
\end{align*}

and the elimination rule,
\[
  \inferrule{\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
             \dots \\
             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
  }{\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
\]

@@

When defining a derivation from $NFA(g \oplus g')$ to $g \oplus g'$, we
wish to simultaneously fold over the mutually recursive definitions comprising $NFA(g \oplus g')$. We may think of this
as an eliminator that replaces each state inside of $NFA(g \oplus g')$ with a tail-recursive call that will ultimately
return a term of type $g \oplus g'$. The structure of such a derivation instatiates the above elimination rule as follow,

\[
  \inferrule{
    \Delta \vdash e : NFA(g \oplus g') \\
    y_{s} : \simulsubst {(X_{s}' \oplus Z'_{s})} \gamma \vdash p_{s} : g \oplus g' \\
    x_{s}' : \simulsubst {(\simulsubst {X_{s}} {\delta_{X}})}{\gamma} \vdash p_{x,s} : g \\
    x_{j}' : \simulsubst {(\simulsubst {X_{j}} {\delta_{X}})}{\gamma} \vdash p_{x,j} : X_{j} \text{~where } j \not \in \{s, e\} \\
    x_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{x, e} : X_{e} \\
    z_{s}' : \simulsubst {(\simulsubst {Z_{s}} {\delta_{Z}})}{\gamma} \vdash p_{z,s} : g' \\
    z_{j}' : \simulsubst {(\simulsubst {Z_{j}} {\delta_{Z}})}{\gamma} \vdash p_{z,j} : Z_{j} \text{~where } j \not \in \{s, e\} \\
    z_{e}' : \simulsubst {Y_{e}}{\gamma} \vdash p_{z,e} : Z_{e} \\
    y_{e} : \simulsubst I {\gamma} \vdash p_{e} : I
  }{\Delta \vdash \mathsf{mfold}(e) : g \oplus g'}
\]

@@

where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively.
This finshes the construction of a parse transformer from $NFA(g \oplus g')$ to $ g \oplus g'$.
The map in the other direction is given below in a couple stages. First, we show how to turn a parse of $NFA(g)$ into a parse of $NFA(g \oplus g')$.

\[
  \inferrule
  {
    \Gamma ; \Delta \vdash u : NFA(g) = \mu(X_{j} = \hat{X}_{j}).X_{s} \\
    \Gamma ; x_{s} : \simulsubst {X_{s}}{\gamma} = X_{s}' \vdash \mathsf{cons}(\textbf{in}_{X_{s}'}(x_{s})) : \mu(\dots).Y_{s} = NFA(g \oplus g')  \\
    \Gamma ; x_{j} : \simulsubst {X_{j}}{\gamma} = X_{j}' \vdash x_{j} : X_{j}', j \not \in \{s, e\} \\
    \Gamma ; x_{e} : \simulsubst {X_{e}}{\gamma} = I \vdash x_{e} : I = X_{e}'
  }
  {\Gamma ; \Delta \vdash \mathsf{mfold}(u) : NFA(g \oplus g')}
\]
where the dots in the above derivation take on the mutually recursive defintions present in $NFA(g \oplus g')$. \steven{Need a better way to display this without blowing up the definitions everywhere} Denote the above embedding of $NFA(g)$ into $NFA(g \oplus g')$ via $\alpha(u) := \mathsf{mfold}(u) : NFA(g \oplus g')$. Similarly define an embedding $\beta$ for an embedding of $NFA(g')$. We can then define the derivation from $g \oplus g'$ to $NFA(g \oplus g')$ by composing the maps isomorphisms $g \overset{\phi}{\cong} NFA(g)$ and $g \overset{\psi}{\cong} NFA(g')$ with $\alpha$ and $\beta$, respectively.

\[
  \inferrule
  {
    p : g \oplus g' \vdash p : g \oplus g' \\
    p_{1} : g \vdash \alpha(\phi(p_{1})) : NFA(g \oplus g') \\
    p_{2} : g' \vdash \beta(\psi(p_{2})) : NFA(g \oplus g')}
  {
    p : g \oplus g' \vdash \textbf{case}~p \{
    \textbf{in}_{g} p_{1} \mapsto \alpha(\phi(p_{1})), \\
    \textbf{in}_{g'} p_{2} \mapsto \beta(\psi(p_{2})) \} : NFA(g \oplus g')}
\]
